Exploring probability distributions for bivariate temporal granularities

### 2020-02-03 main.Rmd
cricket data analysis - code optimised

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others. However, irrespective of the type of cyclic granularities, it can be interesting to visualize the periodicity of  scores/performance under similar data structure as temporal datasets.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. There are many interesting questions that can possibly be answered with such a data set, however, we will explore a few and understand how the proposed approach in the paper can help answer some of the questions.

@@

<!-- The hierarchy table would look like the following: -->

<!-- # ```{r hierarchy2} -->
<!-- # library(gravitas) -->
<!-- # library(tibble) -->
<!-- # hierarchy_model <- tibble::tibble( -->
<!-- #   G = c("over", "inning", "match", "season"), -->
<!-- #   C = c("over-of-inning", "inning-of-match", "match-of-season",1), -->
<!-- #   P = c(20, 2, "aperiodic", 1)) -->
<!-- # knitr::kable(hierarchy_model, format = "markdown") -->
<!-- #  -->
<!-- # ``` -->

@@

Q1: How run rates vary depending on if a team bats first or second?

Mumbai Indians (MI) and Chennai Super kings (CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into this question.

<!-- - Q1: How their run rates vary depending on if they bat first or 2nd? Is there a chance that they are more likely to win if they bat first? -->

@@

From Figure \ref{fig:cricex}, it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that palyers are vulnerable to score through each over and make more runs in the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. Also longer and more distinct letter values in the second innings suggests that the variability is more in the second innings. To answer Q1, winning teams like CSK and MI have indeed different startegies when it comes to batting in the first or second innings.

@@

@@

@@

cricket_fielding <- cricket_season %>% 
  mutate(fielding_proxy = if_else(dismissal_kind %in% c("caught", "caught and bowled"),1,0)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            fielding_wckts = sum(fielding_proxy)) %>%
            mutate(diff_run_rate = c(0,diff(run_rate)))

cricket_tsibble_fld <- cricket_fielding %>% 
  ungroup() %>% 
  mutate(data_index = row_number()) %>% 
  as_tsibble(index = data_index)

cricket_tsibble_fld %>% gran_obs("over", "fielding_wckts")

cricket_tsibble_fld %>% 
  mutate(field = if_else(fielding_wckts==0, "no wickets", ">0 wickets")) %>% 
  #filter(fielding_wckts %in% c(0,1)) %>% 
  prob_plot( "field","over",
                               hierarchy_model,
                               response = "diff_run_rate",
                               plot_type = "boxplot",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("")

@@

```{r crickread, ech0=FALSE}
cricket_tsibble <- gravitas::cricket %>%
  mutate(data_index = row_number()) %>%
  as_tsibble(index = data_index)
```

@@

cricket_bowling <- cricket_season %>% 
  mutate(dot_ball_proxy = if_else(total_runs==0,1,0),
         wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            dot_balls = sum(dot_ball_proxy)) %>% 
            mutate(diff_run_rate = c(0,diff(run_rate)))

cricket_tsibble_bow <- cricket_bowling %>% 
  ungroup() %>% 
  mutate(data_index = row_number()) %>% 
  as_tsibble(index = data_index)

cricket_tsibble_bow %>%
  mutate(dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls")) %>% 
  #filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("dot",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
            symmetric = FALSE)  + ggtitle("")

>>>>

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others. However, irrespective of the type of cyclic granularities, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a  non-temporal dataset similar to any temporal dataset.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. The hierarchy table would look like the following:

```{r hierarchy2}
 library(gravitas)
 library(tibble)
 hierarchy_model <- tibble::tibble(
   G = c("over", "inning", "match", "season"),
   C = c("over-of-inning", "inning-of-match", "match-of-season",1),
   P = c(20, 2, "aperiodic", 1))
 knitr::kable(hierarchy_model, format = "markdown")
```

There are many interesting questions that can possibly be answered with such a data set, however, we will explore a few and understand how the proposed approach in the paper can help answer some of the questions.

@@

@@

Q1: How total runs vary depending on if a team bats first or second?

Mumbai Indians (MI) and Chennai Super kings (CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into this question. Circular cyclic granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circulargran} with over as index of the tsibble. From Figure \ref{fig:cricex}, it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that players are vulnerable to score through each over and make more runs in the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. To answer Q1, winning teams like CSK and MI have indeed different startegies when it comes to batting in the first or second innings.

<!-- Also longer and more distinct letter values in the second innings suggests that the variability is more in the second innings. -->
<!-- - Q1: How their run rates vary depending on if they bat first or 2nd? Is there a chance that they are more likely to win if they bat first? -->

@@

@@

hierarchy_model <- tibble::tibble(
  units = c("index", "over", "inning", "match"),
  convert_fct = c(1, 20, 2, 1)
)

# cricket_tsibble %>%
#   filter(batting_team %in% c("Mumbai Indians",
#                              "Chennai Super Kings"))%>%
#   create_gran("inning_match", hierarchy_model) %>% create_gran("over_inning", hierarchy_model)

cricket_tsibble %>%
  filter(batting_team %in% c("Mumbai Indians",
                             "Chennai Super Kings"))%>%
  gran_advice("inning", 
            "over")

@@

cricket_tsibble %>%
  filter(batting_team %in% c("Mumbai Indians",
                             "Chennai Super Kings"))%>%
  gran_advice("inning", 
            "over")

@@

cricket_dot_field <- cricket_season %>% 
  mutate(fielding_proxy = if_else(dismissal_kind %in% 
                                    c("caught", "caught and bowled"),1,0),
         dot_ball_proxy = if_else(total_runs==0,1,0),
         wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>% 
  group_by(season,
           match_id,
           batting_team,
           bowling_team, 
           inning,
           over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            fielding_wckts = sum(fielding_proxy),
            dot_balls = sum(dot_ball_proxy)) %>%
            mutate(diff_run_rate = c(0,diff(run_rate)))

cricket_tsibble <- cricket_dot_field %>% 
  ungroup() %>% 
  mutate(data_index = row_number()) %>% 
  as_tsibble(index = data_index)

cricket_tsibble %>% gran_obs("over", "fielding_wckts")

cricket_data_ts <- cricket_tsibble %>% 
  mutate(field = if_else(fielding_wckts==0, "no wickets", ">0 wickets"),
         dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls"))

  #filter(fielding_wckts %in% c(0,1)) %>% 
  cricket_data_ts %>% prob_plot( "over","field",
                               hierarchy_model,
                               response = "diff_run_rate",
                               plot_type = "violin",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("")
```

@@

<!-- # ```{r crickread, ech0=FALSE} -->
<!-- # cricket_tsibble <- gravitas::cricket %>% -->
<!-- #   mutate(data_index = row_number()) %>% -->
<!-- #   as_tsibble(index = data_index) -->
<!-- # ``` -->

@@

# library(gravitas)
# library(tsibble)
# 
# 
# cricket_bowling <- cricket_season %>% 
#   mutate(dot_ball_proxy = if_else(total_runs==0,1,0),
#          wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>% 
#   group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
#   summarise(runs_per_over = sum(total_runs),
#             run_rate = sum(total_runs)/length(total_runs),
#             dot_balls = sum(dot_ball_proxy)) %>% 
#             mutate(diff_run_rate = c(0,diff(run_rate)))
# 
# 
# cricket_tsibble_bow <- cricket_bowling %>% 
#   ungroup() %>% 
#   mutate(data_index = row_number()) %>% 
#   as_tsibble(index = data_index)
#

# cricket_tsibble_bow %>%
#   mutate(dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls")) %>% 
#   #filter(dot_balls %in% c(0, 1, 2)) %>%
#   prob_plot("dot",
#             "over",
#             hierarchy_model,
#             response = "diff_run_rate",
#             plot_type = "quantile",
#             quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
#             symmetric = FALSE)  + ggtitle("")

cricket_data_ts %>% 
  #filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("dot",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
            symmetric = FALSE)  + ggtitle("")
```
